---
title: "Predicting Loan Repayment"
output: github_document
date: "2022-12-19"
---
### About the Study and the Data Set
In the lending industry, investors loan money to borrowers with the expectation of receiving repayment with interest. If the borrower repays the loan as agreed, the lender earns a profit from the interest. However, if the borrower is unable to repay the loan, the lender incurs a loss. To mitigate this risk, lenders need to be able to predict the likelihood of a borrower being unable to repay a loan. To help address this issue, we will use a dataset from LendingClub.com, a platform that connects borrowers and investors online. The dataset consists of 9,578 3-year loans made through LendingClub.com from May 2007 to February 2010. The dependent variable "not.fully.paid" indicates whether the loan was not fully repaid, either because the borrower defaulted or the loan was charged off due to the borrower being unlikely to repay it.  

The data set contains the following variables:  

* credit.policy: A binary variable indicating whether the customer meets the credit underwriting criteria of LendingClub.com (1) or not (0).  

* purpose: The reason for taking out the loan, with categories including "credit_card", "debt_consolidation", "educational", "major_purchase", "small_business", and "all_other".  

* int.rate: The interest rate of the loan as a decimal (e.g. a rate of 11% would be stored as 0.11). LendingClub.com tends to assign higher interest rates to borrowers considered to be more risky.  
* installment: The monthly payment amount owed by the borrower if the loan is funded.  
* log.annual.inc: The natural log of the borrower's self-reported annual income.  
* dti: The debt-to-income ratio of the borrower, calculated as the amount of debt divided by annual income.  
* fico: The FICO credit score of the borrower.  
* days.with.cr.line: The number of days the borrower has had a credit line.  
* revol.bal: The borrower's revolving balance, or the amount unpaid at the end of the credit card billing cycle.  
* revol.util: The borrower's utilization rate of their credit line, or the amount of the credit line used relative to the total credit available.  
* inq.last.6mths: The number of inquiries by creditors made to the borrower in the last 6 months.  
* delinq.2yrs: The number of times the borrower has been 30+ days past due on a payment in the past 2 years.  
* pub.rec: The number of derogatory public records the borrower has, such as bankruptcy filings, tax liens, or judgments.  

### EDA
```{r}
loans = read.csv("loans.csv")
str(loans)
```
```{r}
library(knitr)
kable(table(loans$not.fully.paid))
barplot(table(loans$not.fully.paid))

```


Now that we have an idea about the distribution of the loan repayment variable, let's examine the missing data. 
```{r}
kable(summary(loans))
```
We can see that, the variables with missing data are log.annual.inc, days.with.cr.line, revol.util, inq.last.6mths, delinq.2yrs, and pub.rec.  

Lets further investigate the missing records to decide what to do,
```{r}
missing = subset(loans, is.na(log.annual.inc) | is.na(days.with.cr.line) | is.na(revol.util) | is.na(inq.last.6mths) | is.na(delinq.2yrs) | is.na(pub.rec))
nrow(missing)
kable(table(missing$not.fully.paid))
```
Although the number of the missing data is only 62 records, we would want to fill the missing value's instead of ignoring these record so that the model could perhaps cover the cases where some value's does not exist. 

Let's impute the missing data. 
```{r}
library(mice)
#Setting Seeds so that the result is reproducible
set.seed(123)
vars_for_imputation = setdiff(names(loans), "not.fully.paid")
imputed = complete(mice(loans[vars_for_imputation]))
loans[vars_for_imputation] = imputed
```
```{r}
kable(summary(loans))
```

### Building Logistic Regression Model

```{r}
library(caTools)
set.seed(123)
spl = sample.split(loans$not.fully.paid, 0.7)
train = subset(loans, spl == TRUE)
test = subset(loans, spl == FALSE)
model = glm(not.fully.paid ~., data = train, family = binomial)
summary(model)
```
We can see that there are 10 significant variables in our model. 

### Applying the Model
```{r}
test$predicted = predict(model, newdata = test, type = "response")
x = table(test$not.fully.paid, test$predicted >= 0.5)
kable(x)
```

Now, let's calculate the accuracy of the model. 
```{r}
sum(diag(x))/sum(x)
```

Let's see what a base line model accuracy will look like.   
```{r}
x = table(test$not.fully.paid)
kable(x)
```
A simple base line model that assumes a loan will be fully paid all the time (not.fully.paid = 0) will have an accuracy of: 

```{r}
x[1]/sum(x)
```
Lets calculate the AUC of the Model

```{r}
library(ROCR)
pred = prediction(test$predicted, test$not.fully.paid)
as.numeric(performance(pred, "auc")@y.values)
```

### A Possibly Better Base Line Model

LendingClub.com uses the interest rate of a loan as an indicator of the risk associated with that loan. The interest rate is an independent variable in our dataset, and we will use it to compare the risk of different loans and determine whether it can be used as a baseline for ranking loans according to their risk.

```{r}
Model1 = glm(not.fully.paid ~ int.rate, family = binomial, data = train)
summary(Model1)
```
As we can see, the variable int.rate is really significant in the new model, yet it is not significant in a model that contains many variables. This is common when the independent variables are highly correlated, so let's investigate this hypothesis and see. 
```{r}
cor(train$fico, train$int.rate)
```
We can see that there's a strong linear relation between the credit score and the interest rate, the reason it is negative is because ,logically, lower credit score leads to higher interest rate.

```{r}
predModel1 = predict(Model1, newdata = test, type = "response")
summary(predModel1)

```
We can see that the maximum prediction is 0.44 meaning a 0.5 threshold will lead to all the not.fully.paid being 0, meaning all loans will be re-payed.

Let's see the AUC of this Model

```{r}
pred1 = prediction(predModel1, test$not.fully.paid)
as.numeric(performance(pred1, "auc")@y.values)
```
So given that this model has a not so different AUC from the previous one, while being way simpler given that it have only one independent variable, it can be considered a better base line model. 

### Different Investment Stratgies 

The goal of an investor is to find loans that are likely to be profitable, meaning that they will be paid back in full and the investor will earn interest on the loan. However, there is also a risk that the loan will not be paid back, in which case the investor will lose the money invested. Therefore, the investor should seek loans that offer a good balance between risk and reward, meaning they have a reasonable chance of being paid back while still offering a potentially high return on investment.  

First, let's establish some ideas:   

* $c*exp(rt)$ is the interest revenue given that $c$ is the initial investment, $r$ is the interest rate, and $t$ is the time.  

* $c*exp(rt)-c$ is the revenue of the said investment after deducting the initial investment capital given that the loan has been fully repaid.  

* $-c$ is the maximum loss an investor could take in case the loan wasn't repaid at all.    


***To evaluate the effectiveness of an investment strategy, we need to calculate the potential profit for each loan in the test set. We will assume that the investor has invested \$1 (c=1) in each loan. To calculate the profit for a fully paid loan, we will use the formula exp(rt)-1, where r is the interest rate and t is the length of the loan in years. In this case, all of the loans in the dataset are 3-year loans, so t=3 in our calculations. We will assign this value to every observation and then replace it with -1 for any loans that were not paid in full. This will allow us to compare the potential profits for different loans and evaluate the quality of the investment strategy.***

```{r}
test$profit = exp(test$int.rate*3)-1
test$profit[test$not.fully.paid == 1] = -1 
```

```{r}
summary(test$profit)
```
An investment strategy of equally investing in all loans would yield a profit of \$20.94 for a \$100 investment, but this approach does not take advantage of the prediction model that was developed earlier. Investors generally want to find loans that offer a good balance between reward (high interest rates) and risk (low likelihood of not being paid back).

To meet this goal, we will examine an investment strategy that involves only investing in loans with high interest rates (at least 15%) and selecting the ones with the lowest predicted risk of not being fully paid back. We will model an investor who invests $1 in each of the 100 most promising loans. This strategy aims to maximize the potential return on investment while minimizing the risk of loss.

```{r}
highinterest = subset(test, int.rate >= 0.15)
mean(highinterest$profit)
```
```{r}
x = table(highinterest$not.fully.paid)
kable(x)
```
```{r}
x[2]/sum(x)
```
We can see that almost 25% of the loans with high interest rate didn't get paid in full, and the mean of profit for the whole is 22%.  

To identify the 100 loans with the lowest predicted risk of not being paid back in full, we will first sort the predicted risks in increasing order. Then, we will select the 100th element of this sorted list, which will be the 100th smallest predicted probability of not paying in full. This will allow us to identify the 100 loans with the lowest predicted risk and consider them for investment.  
```{r}
cutmark = sort(highinterest$predicted, decreasing = FALSE)[100]
selected_investments = subset(highinterest, predicted <= cutmark)

```

Now let's see the result of the strategy

```{r}
nrow(selected_investments)
mean(selected_investments$profit)
```
```{r}
x = table(selected_investments$not.fully.paid)
kable(x)
```
```{r}
x[2]/sum(x)
```
***As we can see, even though the model we have built is simple, coupled with an investment strategy, it became quite effective given us a profit rate of almost 39% and a loan repayment rate of 86%. This is a great improvement from 22% and 75%, respectively.***
 







